---
title: "Evironmental data download and cleaning"
author: "Florencia Grattarola"
date: "2021-08-14"
output: 
  html_document: 
    keep_md: yes
    toc: yes
    toc_float: yes
    highlight: pygments
    theme: flatly
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dpi=200, echo = TRUE, message = FALSE, warning = FALSE)
```

```{r, message = F, warning=F}
library(rnaturalearth)
library(tmap)
library(stars)
library(terra)
terraOptions(tempdir = 'big_data/temp')
library(sf)
sf::sf_use_s2(FALSE)
library(tidyverse)
```

# Basemaps

```{r latam, eval=T, warning=F}
# world and Latin America
world <- rnaturalearth::ne_countries(scale = 'large', returnclass = 'sf') %>% st_set_crs("EPSG:4326")
bbox_Latam_unprojected <- c(xmin=-118.40137, ymin=-55.89170, xmax=-34.80547, ymax= 32.71533)
Latam_unprojected <- world %>% st_crop(bbox_Latam_unprojected) %>% st_make_valid

# equal area projection (Equatorial Lambert azimuthal equal-area) 
equalareaCRS <-  '+proj=laea +lon_0=-73.125 +lat_0=0 +datum=WGS84 +units=m +no_defs'
Latam_projected <- sf::st_transform(Latam_unprojected, crs=equalareaCRS) %>% st_cast() %>% st_make_valid
Latam_projected <- Latam_projected %>% 
  mutate(iso_a2=ifelse(is.na(iso_a2) & name=='France', 'GF', iso_a2)) %>% 
  mutate(name_en=ifelse(iso_a2=='GF' & name_en=='France', 'French Guiana', name_en))

Latam <- st_union(Latam_projected) 
```

```{r countryraster, eval=T, warning=F}
Latam.grid <- st_make_grid(Latam+1000, cellsize = 100000, crs = equalareaCRS) %>% st_sf('geometry' = ., 'occ'= 0)
r <- rast(res=100000, ext(vect(Latam.grid)), crs=equalareaCRS)
Latam.raster <- terra::rasterize(x = vect(Latam.grid), y = r, field = 'occ') %>% terra::mask(., vect(Latam))

# raster of country names
Latam.grid_countries <- st_join(Latam.grid, Latam_projected %>% dplyr::select(iso_a2))

Latam_countries <- terra::rasterize(x = vect(Latam.grid_countries),
                 y = Latam.raster,
                 field = 'iso_a2',
                 fun = 'first') %>% terra::mask(., vect(Latam))
names(Latam_countries) <- 'country'

#plot(Latam_countries)

tmap::tm_shape(Latam_countries) +
    tmap::tm_raster(col='country', style='cat', palette = "Set1")+
    tm_layout(legend.show = F) +
    tmap_options(max.categories = 50)
```

# Data Sources
## Download functions

```{r, eval=F}
# download functions
download <- function(download_files, download_url){
  for(file_i in download_files){
    url <- str_c(download_url, file_i)
    download.file(url, destfile=str_c('big_data/', file_i))
    if (endsWith(file_i, '.zip')) {
      unzip(str_c('big_data/', file_i), exdir=str_c('big_data/', str_remove(file_i, '.zip')))
      if (file.exists(str_c('big_data/', file_i))) {
        file.remove(str_c('big_data/', file_i))
      }
    }
  }
}
download_worldPop <- function(web_url, years, resolution) {
  for(year in years) {
    url <- str_glue('{web_url}{year}/0_Mosaicked/ppp_{year}_{resolution}_Aggregated.tif')
    path <- 'big_data/popu/'
    download.file(url, destfile = str_glue('{path}/ppp_{year}_{resolution}_Aggregated.tif'))
  }
}

# extend the timeout for downloads
getOption('timeout') 
options(timeout=500) 
```

## Environmental data
SOURCE: WorldClim V2.1
Data download at 30 seconds resolution 

```{r, eval=F}
# download_files <- c('wc2.1_30s_elev.zip', 'wc2.1_30s_tavg.zip', 'wc2.1_30s_prec.zip')
# download_url <- 'https://biogeo.ucdavis.edu/data/worldclim/v2.1/base/'
# download(download_files, download_url)

# elevation
world_elev <- rast('big_data/wc2.1_30s_elev/wc2.1_30s_elev.tif')
elev_unprojected <- crop(world_elev, ext(Latam_unprojected)+1) 
elev <- project(elev_unprojected, equalareaCRS, method='bilinear') # reproject to equal area
names(elev) <- 'elev'
rm(elev_unprojected)

# average annual temperature
tavg_files <- list.files('big_data/wc2.1_30s_tavg', '.tif', full.names=TRUE)
world_tavg <- rast(tavg_files)
tavg_unprojected <- crop(world_tavg, ext(Latam_unprojected)+1)
tavg <- project(tavg_unprojected, equalareaCRS, method='bilinear') # reproject to equal area
tavg <- mean(tavg)
names(tavg) <- 'tavg'
rm(tavg_unprojected)

# annual precipitation
prec_files <- list.files('big_data/wc2.1_30s_prec', '.tif', full.names=TRUE)
world_prec <- rast(prec_files)
prec_unprojected <- crop(world_prec, ext(Latam_unprojected)+1)
prec <- project(prec_unprojected, equalareaCRS, method='bilinear') # reproject to equal area
prec <- mean(prec)
names(prec) <- 'prec'
rm(prec_unprojected)
```

```{r elev, echo=F, warning=F, message=F}
elev <- rast('big_data/elev.tif')
tmap::tmap_mode(mode ='view')

tmap::tm_shape(elev) +
    tmap::tm_raster(palette = 'BrBG', midpoint = NA, style= "cont", n=10) +
    tmap::tm_legend(position = c('left', 'bottom'))

```

```{r, eval=F}

# bioclimatic variables
bio_files <- list.files('big_data/wc2', '.tif', full.names = T)
world_bio <- rast(bio_files)
bio_unprojected <- crop(world_bio, ext(Latam_unprojected)+1)
bio <- project(bio_unprojected, equalareaCRS, method='bilinear') # reproject to equal area
names(bio) <- sprintf('bio_%s',seq(1:19))
rm(bio_unprojected)

# BIO1 = Annual Mean Temperature
# BIO2 = Mean Diurnal Range (Mean of monthly (max temp - min temp))
# BIO3 = Isothermality (BIO2/BIO7) (×100)
# BIO4 = Temperature Seasonality (standard deviation ×100)
# BIO5 = Max Temperature of Warmest Month
# BIO6 = Min Temperature of Coldest Month
# BIO7 = Temperature Annual Range (BIO5-BIO6)
# BIO8 = Mean Temperature of Wettest Quarter
# BIO9 = Mean Temperature of Driest Quarter
# BIO10 = Mean Temperature of Warmest Quarter
# BIO11 = Mean Temperature of Coldest Quarter
# BIO12 = Annual Precipitation
# BIO13 = Precipitation of Wettest Month
# BIO14 = Precipitation of Driest Month
# BIO15 = Precipitation Seasonality (Coefficient of Variation)
# BIO16 = Precipitation of Wettest Quarter
# BIO17 = Precipitation of Driest Quarter
# BIO18 = Precipitation of Warmest Quarter
# BIO19 = Precipitation of Coldest Quarter
```

```{r bio, echo=F, warning=F, message=F}
bio <- rast('big_data/bio.tif')

# rasterVis::gplot(bio) + geom_tile(aes(fill = value)) +
#   scale_fill_distiller(palette = 'BrBG' , na.value = NA) +
#   labs(fill='elevation', x='', y='') +
#   coord_equal() +
#   theme_bw()
plot(bio)
```

## Land cover
SOURCE: MODIS Land Cover Type Yearly L3 Global 500m

```{r land.download, eval=F}
MODIStsp(gui             = FALSE,
         out_folder      = 'big_data/',
         out_folder_mod  = 'big_data/',
         selprod         = 'LandCover_Type_Yearly_500m (MCD12Q1)',
         bandsel         = 'LC1', 
         sensor          = 'Terra',
         user            = '' , # your username for NASA http server
         password        = '',  # your password for NASA http server
         start_date      = '2000.01.01', 
         end_date        = '2022.12.31', 
         verbose         = TRUE,
         bbox            =  c(-5596641.0845, -6673508.6914, 4698677.0087, 4157242.8202), #bbox of Latam
         spatmeth        = 'bbox',
         out_format      = 'GTiff',
         compress        = 'LZW',
         out_projsel     = 'User Defined',
         output_proj     = '+proj=laea +lon_0=-73.125 +lat_0=0 +datum=WGS84 +units=m +no_defs',
         delete_hdf      = TRUE,
         parallel        = TRUE
         )
```

``` {r land.processing, eval=F}
landcover_pre_files <- list.files('big_data/LandCover_Type_Yearly_500m_v6/LC1', '200[0-1]|201[0-3]', full.names = T)
landcover_pre_c <- rast(landcover_pre_files)
landcover_pre <- modal(landcover_pre_c) %>% mask(., vect(Latam))
names(landcover_pre) <- 'landcover_pre'
rm(landcover_pre_c)

landcover_pos_files <- list.files('big_data/LandCover_Type_Yearly_500m_v6/LC1', '201[4-9]|202[0-9]', full.names = T)
landcover_pos_c <- rast(landcover_pos_files)
landcover_pos <- modal(landcover_pos_c) %>% mask(., vect(Latam))
names(landcover_pos) <- 'landcover_pos'
rm(landcover_pos_c)
```

### Example plot: Land at pre period 
```{r land.plot, echo=F, warning=F, message=F, eval=F}
land_pre <- rast('big_data/land_pre.tif')
land_pre <- as.factor(land_pre$landcover_pre)

levels(land_pre) <- c("Evergreen needleleaf forests",
                      "Evergreen broadleaf forests",
                      "Deciduous needleleaf forests",
                      "Deciduous broadleaf forests",
                      "Mixed forests",
                      "Closed shrublands",
                      "Open shrublands",
                      "Woody savannas",
                      "Savannas",
                      "Grasslands",
                      "Permanent wetlands",
                      "Croplands",
                      "Urban and built-up lands",
                      "Cropland/natural vegetation mosaics",
                      "Snow and ice",
                      "Barren",
                      "Water bodies")

tmap::tmap_mode(mode ='view')
tmap::tm_shape(land_pre) +
    tmap::tm_raster(palette = 'Set2', style= "cat", labels = levels(land_pre)[[1]])
```

## Vegetation Continuous Fields
SOURCE: MODIS Vegetation Continous Fields Yearly L3 Global 250m

```{r veg.download, eval=F}
MODIStsp(gui             = FALSE,
         out_folder      = 'big_data/',
         out_folder_mod  = 'big_data/',
         selprod         = 'Veg_Cont_Fields_Yearly_250m (MOD44B)',
         bandsel         = c('Percent_Tree_Cover', 'Percent_NonTree_Vegetation', 'Percent_NonVegetated'), 
         sensor          = 'Terra',
         user            = '' , # your username for NASA http server
         password        = '',  # your password for NASA http server
         start_date      = '2000.01.01', 
         end_date        = '2022.12.31', 
         verbose         = TRUE,
         bbox            =  c(-5596641.0845, -6673508.6914, 4698677.0087, 4157242.8202), #bbox of Latam
         spatmeth        = 'bbox',
         out_format      = 'GTiff',
         compress        = 'LZW',
         out_projsel     = 'User Defined',
         output_proj     = '+proj=laea +lon_0=-73.125 +lat_0=0 +datum=WGS84 +units=m +no_defs',
         delete_hdf      = TRUE,
         parallel        = TRUE,
         reprocess = TRUE
         )
```

``` {r veg.processing, eval=F}
veg_pre_files <- list.files('big_data/Veg_Cont_Fields_Yearly_250m_v6/Perc_TreeCov/', '200[0-9]|201[0-3]', full.names = T)
veg_pre_c <- rast(veg_pre_files)
veg_pre <- mean(veg_pre_c, na.rm=T) %>% mask(., vect(Latam))
names(veg_pre) <- 'veg_pre'
rm(veg_pre_c)

veg_pos_files <- list.files('big_data/Veg_Cont_Fields_Yearly_250m_v6', '201[4-9]|202[0-9]', full.names = T)
veg_pos_c <- rast(veg_pos_files)
veg_pos <- mean(veg_pos_c, na.rm=T) %>% mask(., vect(Latam))
names(veg_pos) <- 'veg_pos'
rm(veg_pos_c)
terra::writeRaster(veg_pos, 'big_data/veg_pos.tif', overwrite=TRUE)
```

### Example plot: Vegetation at pre period 
```{r veg.plot, echo=F, warning=F, message=F}
veg_pre <- rast('big_data/veg_pre.tif')
tmap::tmap_mode(mode ='view')

tmap::tm_shape(veg_pre) +
    tmap::tm_raster(palette = 'YlGn', midpoint = NA, style= "cont")
```

``` {r nontree.processing, eval=F}
nontree_pre_files <- list.files('big_data/Veg_Cont_Fields_Yearly_250m_v6/Perc_TreeCov/', '200[0-9]|201[0-3]', full.names = T)
nontree_pre_c <- rast(nontree_pre_files)
nontree_pre <- mean(nontree_pre_c, na.rm=T) %>% mask(., vect(Latam))
names(nontree_pre) <- 'nontree_pre'
rm(nontree_pre_c)

nontree_pos_files <- list.files('big_data/Veg_Cont_Fields_Yearly_250m_v6', '201[4-9]|202[0-9]', full.names = T)
nontree_pos_c <- rast(nontree_pos_files)
nontree_pos <- mean(nontree_pos_c, na.rm=T) %>% mask(., vect(Latam))
names(nontree_pos) <- 'nontree_pos'
rm(nontree_pos_c)
terra::writeRaster(nontree_pos, 'big_data/nontree_pos.tif', overwrite=TRUE)
```

### Example plot: Non tree vegetation at pre period 
```{r nontree.plot, echo=F, warning=F, message=F}
nontree_pre <- rast('big_data/nontree_pre.tif')
tmap::tmap_mode(mode ='view')

tmap::tm_shape(nontree_pre) +
    tmap::tm_raster(palette = 'YlGn', midpoint = NA, style= "cont")
```


## Net Primary Production (NPP)
SOURCE: MODIS Net Primary Production Yearly L4 Global 1km

```{r npp.download, eval=F}
MODIStsp(gui             = FALSE,
         out_folder      = 'big_data/',
         out_folder_mod  = 'big_data/',
         selprod         = 'Net_PP_GapFil_Yearly_500m (M*D17A3HGF)',
         bandsel         = 'Npp', 
         sensor          = 'Terra',
         user            = '' , # your username for NASA http server
         password        = '',  # your password for NASA http server
         start_date      = '2000.01.01', 
         end_date        = '2022.12.31', 
         verbose         = TRUE,
         bbox            =  c(-5596641.0845, -6673508.6914, 4698677.0087, 4157242.8202), #bbox of Latam
         spatmeth        = 'bbox',
         out_format      = 'GTiff',
         compress        = 'LZW',
         out_projsel     = 'User Defined',
         output_proj     = '+proj=laea +lon_0=-73.125 +lat_0=0 +datum=WGS84 +units=m +no_defs',
         delete_hdf      = TRUE,
         parallel        = TRUE
         )
```

``` {r npp.processing, eval=F}
npp_pre_files <- list.files('big_data/Net_PP_Yearly_500m_v6/Npp', '200[0-9]|201[0-3]', full.names = T)
npp_pre_c <- rast(npp_pre_files)
npp_pre <- mean(npp_pre_c) %>% mask(., vect(Latam))
names(npp_pre) <- 'npp_pre'
rm(npp_pre_c)

npp_pos_files <- list.files('../../../OneDrive - CZU v Praze/big_data/Net_PP_Yearly_500m_v6/Npp', '201[4-9]|202[0-9]', full.names = T)
npp_pos_c <- rast(npp_pos_files)
npp_pos <- mean(npp_pos_c) %>% mask(., vect(Latam))
names(npp_pos) <- 'npp_pos'
rm(npp_pos_c)
```

### Example plot: NPP at pre period 
```{r npp.plot, echo=F, warning=F, message=F}
npp_pre <- rast('big_data/npp_pre.tif')
tmap::tmap_mode(mode ='view')

tmap::tm_shape(npp_pre) +
    tmap::tm_raster(palette = 'BuGn', midpoint = NA, style= "cont")
```


## Thining data
### Accessibility
SOURCE: Weiss et al. (2019) https://doi.org/10.1038/nature25181
vector 1:110m scale - Version 4.1.0

```{r acce.processing, eval=F}
world_acce <- rast('big_data/acc_50k.tif')
acce_unprojected <- crop(world_acce, ext(Latam_unprojected)+1) 
acce <- project(acce_unprojected, equalareaCRS, method='near') # reproject to equal area
names(acce) <- 'acce'
rm(acce_unprojected)
```

```{r acce.plot, echo=T, warning=F, message=F}
acce <- rast('big_data/acce.tif')

tmap::tmap_mode(mode ='view')
tmap::tm_shape(acce) +
    tmap::tm_raster(palette = 'Spectral', n=12, style = 'cont')
```

# Save data

```{r, eval=F}
# covariates
terra::writeRaster(elev, 'big_data/elev.tif', overwrite=TRUE)
terra::writeRaster(tavg, 'big_data/tavg.tif', overwrite=TRUE)
terra::writeRaster(prec, 'big_data/prec.tif', overwrite=TRUE)

terra::writeRaster(bio, 'big_data/bio.tif', overwrite=TRUE)

terra::writeRaster(landcover_pre, 'big_data/land_pre.tif', overwrite=TRUE)
terra::writeRaster(landcover_pos, 'big_data/land_pos.tif', overwrite=TRUE)

terra::writeRaster(veg_pre, 'big_data/veg_pre.tif', overwrite=TRUE)
terra::writeRaster(veg_pos, 'big_data/veg_pos.tif', overwrite=TRUE)

terra::writeRaster(nontree_pre, 'big_data/nontree_pre.tif', overwrite=TRUE)
terra::writeRaster(nontree_pos, 'big_data/nontree_pos.tif', overwrite=TRUE)

terra::writeRaster(npp_pre, 'big_data/npp_pre.tif', overwrite=TRUE)
terra::writeRaster(npp_pos, 'big_data/npp_pos.tif', overwrite=TRUE)

# thinning variables
terra::writeRaster(popu, 'big_data/popu.tif', overwrite=TRUE)
terra::writeRaster(acce, 'big_data/acce.tif', overwrite=TRUE)
```

```{r basemaps}
# basemaps
st_write(Latam, 'data/Latam_vector.shp', append = F)
st_write(Latam_projected, 'data/Latam_vector_countries.shp', append = F)
terra::writeRaster(Latam.raster, 'data/Latam_raster.tif', overwrite=TRUE)
terra::writeRaster(Latam_countries, 'data/Latam_raster_countries.tif', overwrite=TRUE)
```

As a general practice, you should not save `Raster*` objects to rds data files. Use `writeRaster` instead and save as a tif. This is because large objects are backed by temporary files on disk that disappear at the end of a session. – Robert Hijmans 
